---
title: "Data Preparation"
author: "Laura Cosgrove"
date: "5/1/2019"
output: pdf_document
---

```{r}
library(tidyverse)
```

# Introduction to Dataset


The Open Access Series of Imaging Studies (OASIS) is a project whose
aim is to encourage open analysis of neuroimaging datasets in the scientific community
through free distribution and compilation. The OASIS-3 release is a retrospective compilation of longitudinal neuroimaging, clinical, cognitive, and biomarker data of 1098 participants in several research studies at the Charles F. and Joanne Knight Alzheimer's Disease Research Center at Washington University in St. Louis over the course of 30 years. Of included participants, 609 are cognitively normal and 489 are in various stages of cognitive decline, with ages ranging from 42 - 95 years.

First, we'll prepare the data by coding the longitudinal IDs to days and years:


```{r prep data}
freesurfer = read_csv("./data/freesurfer.csv") %>% 
  janitor::clean_names() %>% 
  select(-fs_date, -included_t1s)
clinical_data = read_csv("./data/ClinicalData.csv") %>% 
  janitor::clean_names() %>% 
  select(-date, -age)

#year preparation 
clinical_data <- clinical_data %>% 
  mutate(day = str_extract(adrc_adrcclinicaldata_id, "[^_]+$"),
         day = as.numeric(str_remove(day, "d")),
         year = day/365,
         subject = as.numeric(str_remove(subject, "OAS")))

freesurfer <- freesurfer %>% 
   mutate(day = str_extract(fs_fsdata_id, "[^_]+$"),
         day = as.numeric(str_remove(day, "d")),
         year = day/365,
         subject = as.numeric(str_remove(subject, "OAS"))) 


```

# Exploratory Data Analysis

First, we examine the longitudinal structure of the data. What does the Freesurfer-derived MRI data look like over time when patients are classified based on their highest diagnosed CDR (Clinical Dementia Rating) score? 

```{r get highest-ever CDR scores}
#List of highest-ever CDR score
subjects_ever_dementia <- clinical_data %>% 
  distinct(subject, cdr, year) %>%
  mutate(year = glue::glue("year_{round(year)}")) %>% 
  rowid_to_column() %>% 
  spread(year, cdr) 

year_cols <- subjects_ever_dementia %>% select(year_1:year_9) %>% names()

subjects_ever_dementia <- subjects_ever_dementia %>% 
  mutate_all(~replace(., is.na(.), 0)) %>% 
  mutate(cdr_highest = pmax(!!!rlang::syms(year_cols))) %>% 
  mutate(cdr_highest = factor(cdr_highest)) %>% 
  select(subject, cdr_highest)

```

```{r plot anatomical volumes over time}

#ICV/eTIV
freesurfer %>% 
  left_join(subjects_ever_dementia, by = "subject") %>% 
  ggplot(aes(y = intra_cranial_vol, x = year, group = subject, color = cdr_highest)) +
  facet_wrap(~cdr_highest, nrow = 2) +
  geom_line() + theme_minimal() + scale_color_viridis_d()


#Total gray volume/icv
freesurfer %>% 
  left_join(subjects_ever_dementia, by = "subject") %>% 
  ggplot(aes(y = total_gray_vol/intra_cranial_vol, x = year, group = subject, color = cdr_highest)) +
  facet_wrap(~cdr_highest, nrow = 2) +
  geom_line() + theme_minimal() + scale_color_viridis_d()

#below doesn't look great, maybe b/c of different methods: NOTE: SubCortGray excludes brain stem and cerebellum with 5.2.
freesurfer %>% 
  left_join(subjects_ever_dementia, by = "subject") %>% 
  ggplot(aes(y = sub_cort_gray_vol/intra_cranial_vol, x = year, group = subject, color = cdr_highest)) +
  facet_wrap(~cdr_highest, nrow = 2) +
  geom_line() + theme_minimal() + scale_color_viridis_d()

#total white matter/cv
freesurfer %>% 
  left_join(subjects_ever_dementia, by = "subject") %>% 
  ggplot(aes(y = cortical_white_matter_vol/intra_cranial_vol, x = year, group = subject, color = cdr_highest)) +
  facet_wrap(~cdr_highest, nrow = 2) +
  geom_line() + theme_minimal() + scale_color_viridis_d()

```

# Selecting a time slice

To proceed with our analysis, we'll need to abandon the longitudinal structure of the data. However, we want to preserve as much data as possible while matching anatomical scores to the outcome.

```{r}
clinical_data %>% select(subject, day, year) %>% 
  ggplot(aes(x = year)) +
  geom_histogram()
 
freesurfer %>% select(subject, day, year) %>% 
  ggplot(aes(x = year)) +
  geom_histogram()

#Round to closest year in both datasets 
freesurfer <- freesurfer %>% 
  mutate(year_round = round(year))

#Merge and remove nonmatched values: effectively matching +- 6 mo.
merged_data <- clinical_data %>% 
  mutate(year_round = round(year)) %>% 
  left_join(freesurfer, by = c("subject", "year_round")) %>% 
  filter(!is.na(adrc_adrcclinicaldata_id), !is.na(fs_fsdata_id)) %>% 
  rename(year_clinical = year.x, day_clinical = day.x, 
         year_mri = year.y, day_mri = day.y)

#If two entries per year, take the latest one:
merged_data <- merged_data %>% 
  group_by(subject, year_round) %>% 
  summarize_all(last) %>% 
  ungroup()

#Take baseline
baseline_data <- merged_data %>% 
  group_by(subject) %>% 
  summarize_all(first)

#Take the last measurement 
last_data <- merged_data %>% 
  group_by(subject) %>% 
  summarize_all(last)

#Take the 2nd measurement
second_data <- merged_data %>% 
  group_by(subject) %>% 
  summarize_all(~nth(., 2)) %>% 
  filter(!is.na(year_round))

#Take the 3nd measurement
third_data <- merged_data %>% 
  group_by(subject) %>% 
  summarize_all(~nth(., 3)) %>% 
  filter(!is.na(year_round))

```

# Data preparation 

First, we needed to define a rule for matching clinical measurements and MRI data on time. We chose to match based on the closest year, effectively a +- 6 months of difference. When multiple measurements were taken in a single year, we relied on the latest measurement for simplicity.

Although all patients underwent baseline clinical evaluations, some patients only began MRI screens many years into the study. To avoid constraining our sample size to a large extent, after matching the appropriate clinical measurement to the MRI measurements, we did not slice the data on one specific year from baseline; rather, we took the last measurement available for each subject, in the hopes of capturing a more diverse pool of cognitive decline. This leaves us with data from 997 participants, whose clinical and MRI data are matched on time-point.

First, we will dichotomize the outcome into dementia/non dementia by classifying cdr = 0 as non-dementia, and cdr > 0 as dementia. We will not consider a multi-class classification model, because of the imbalance of classes:

```{r}
last_data %>% 
  mutate(cdr = factor(cdr)) %>% 
  count(cdr) %>% 
  knitr::kable()
```

```{r dichotomizing outcome}
cog_data <- last_data %>% 
  mutate(cdr = factor(case_when(cdr == 0 ~ 0,
                                cdr > 0 ~ 1), 
                      levels = c(0, 1),
                      labels = c("NonDementia", "Dementia"))) %>% 
  select(-adrc_adrcclinicaldata_id, -fs_fsdata_id, -starts_with("dx"), -starts_with("day")) %>% 
  mutate(age = age_at_entry + year_round) %>% 
  select(subject, age, year_round, cdr, mmse, apoe, height, weight, intra_cranial_vol:cortical_white_matter_vol) 
  
```

Next, we will code APOE into counts of protective alleles versus risk alleles:

```{r coding apoe}
#Coding APOE into count of protective allele vs risk allele
cog_data <- cog_data %>% 
  mutate(apoe = as.character(apoe)) %>% 
  separate(apoe, into = c("apoe_one", "apoe_two"), sep = 1) %>% 
  mutate(apoe_one_e2 = if_else(apoe_one == "2", 1, 0),
         apoe_one_e4 = if_else(apoe_one == "4", 1, 0),
         apoe_two_e2 = if_else(apoe_two == "2", 1, 0),
         apoe_two_e4 = if_else(apoe_two == "4", 1, 0)) %>% 
  mutate(protective_e2 = apoe_one_e2 + apoe_two_e2,
        risk_e4 = apoe_one_e4 + apoe_two_e4) %>% 
  select(subject, year_round, cdr, age, mmse, protective_e2, risk_e4, everything(), -starts_with("apoe"))

```

Finally, write our csv:

```{r}
write_csv(cog_data, "./data/cog_data")
saveRDS(cog_data, "./data/cog_data.RDS")
```

## Plots

```{r}
cog_data %>% 
  gather(anatomical_volumes, volume, intra_cranial_vol:cortical_white_matter_vol) %>% 
  ggplot(aes(x = anatomical_volumes, y = volume, fill = cdr)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() + scale_fill_viridis_d()
```

```{r}
cog_data %>% 
  gather(allele, count, protective_e2, risk_e4) %>% 
  group_by(cdr, count) %>% 
  count(allele) %>% 
  drop_na() %>% 
  mutate(allele_count = str_c(allele, count, sep = "_")) %>% 
  ggplot(aes(x = allele_count, y = n, fill = cdr)) +
  geom_col() +
  coord_flip() +
  facet_grid(~cdr) + scale_fill_viridis_d() + theme_minimal()
  

#Relatively equal distributions, slightly more risk alleles in dementia
table(cog_data$cdr, cog_data$protective_e2)

table(cog_data$cdr, cog_data$risk_e4)

```

From this, it appears that the presence of risk alleles may be more important that the protective alleles as a risk factor for dementia. Further, we note small differences in cortical volumes (aside from total intracranial volume, which is a corrective factor not strictly of interest): white matter volume, cortex volume, and grey volume seem slightly smaller in the dementia group.

#kmeans 

```{r}
library(factoextra) #provides visualization tools for clustering and PCA (ggplot based)
library(gridExtra)
library(corrplot)
library(RColorBrewer)  #nice color palettes
library(gplots) #heatmap for better visualization
library(caret)

cog_data <- readRDS("./data/cog_data_preproc.RDS")

set.seed(1)
train_index <- createDataPartition(cog_data$cdr, p = 2/3, list = FALSE, times = 1)
cog_train <- cog_data[train_index,] 

set.seed(12)
preProc_fn <- preProcess(cog_train[3:10], 
           method = c("center", "scale", "knnImpute"),
          k = 5,
          knnSummary = mean,
          verbose = TRUE)

cog_data[3:10] <- predict(preProc_fn, cog_data[3:10]) 

# select k through silhouette
fviz_nbclust(cog_data[3:10],
             FUNcluster = kmeans,
             method = "silhouette")

set.seed(1) #because there is randomness in the results
km <- kmeans(cog_data[3:10] , centers = 3, nstart = 20) #because the previous results gave us 3 as the best K; nstart = 20 so multiple random starts

km_vis <- fviz_cluster(list(data = cog_data[3:10], cluster = km$cluster), 
                        ellipse.type = "convex", geom = "point", 
                        palette = "Dark2") + labs(title = "K-means") 

km_vis

cog_data %>% 
  mutate(cluster_membership = factor(km$cluster)) %>% 
  group_by(cluster_membership) %>% 
  count(cdr) %>% 
  spread(cdr, n) %>% 
  mutate('Cluster Percent Dementia' = Dementia/(NonDementia+ Dementia))

#select k through gap

fviz_nbclust(cog_data[3:10],
             FUNcluster = kmeans,
             method = "gap")

set.seed(1) #because there is randomness in the results
km_2 <- kmeans(cog_data[3:10] , centers = 6, nstart = 20) 

km_vis <- fviz_cluster(list(data = cog_data[3:10], cluster = km$cluster), 
                        ellipse.type = "convex", geom = "point", 
                        palette = "Dark2") + labs(title = "K-means") 

km_vis

cog_data %>% 
  mutate(cluster_membership = factor(km_2$cluster)) %>% 
  group_by(cluster_membership) %>% 
  count(cdr) %>% 
  spread(cdr, n) %>% 
  mutate('Cluster Percent Dementia' = Dementia/(NonDementia + Dementia))

#Plots on brain volume show the clustering algorithm found variability not explained by dementia status.
cog_data %>% 
   mutate(cluster_membership = factor(km_2$cluster)) %>% 
  ggplot(aes(y = rh_cortical_white_matter_vol, x = cluster_membership, fill = cdr)) +
  coord_flip() +
  geom_boxplot() + theme_bw() + scale_fill_viridis_d()

cog_data %>% 
   mutate(cluster_membership = factor(km_2$cluster)) %>% 
  ggplot(aes(y = lh_cortex_vol, x = cluster_membership, fill = cdr)) +
  coord_flip() +
  geom_boxplot() + theme_bw() + scale_fill_viridis_d()

cog_data %>% 
   mutate(cluster_membership = factor(km_2$cluster)) %>% 
  ggplot(aes(y = intra_cranial_vol, x = cluster_membership, fill = cdr)) +
  coord_flip() +
  geom_boxplot() + theme_bw() + scale_fill_viridis_d()

#not terribly useful plot of cluster means of standardized counts of e2/e4 alleles  by dementia status
cog_data %>% 
   mutate(cluster_membership = factor(km_2$cluster)) %>% 
  group_by(cluster_membership, cdr) %>% 
  count(mean_risk = mean(risk_e4), mean_protective = mean(protective_e2)) %>% 
  ggplot(aes(x = mean_risk, y = mean_protective, size = n, color = cdr)) +
  geom_point(alpha = 0.9) + theme_bw()

```

